---
title: "Cauda EVs miRNAs analysis using data from both runs"
author: "Anar Alshanbayeva & Deepak Tanwar"
date: "`r Sys.Date()`"
output:
  rmdformats::readthedown:
    highlight: tango
    self_contained: true
    thumbnails: true
    lightbox: true
    gallery: false
    keep_md: no
    number_sections: no
    fig_width: 8
    fig_height: 8
    fig_caption: true
    df_print: paged
    code_folding: show
  fontsize: 12pt
  geometry: margin=1in
  documentclass: article
# bibliography: references.bib
link-citations: yes
---


```{r, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

# Libraries required for data analysis
```{r, message=FALSE, warning=FALSE}
source("src.R")
```


# Data: read counts table from both runs
```{r}
counts_total_run3 <- read.table(
  file = "run3_excerpt_counts/exceRpt_miRNA_ReadCounts.txt",
  header = TRUE, sep = "", row.names = 1
)

counts_total_run4 <- read.table(
  file = "run4_excerpt_counts_/exceRpt_miRNA_ReadCounts.txt",
  header = TRUE, sep = "", row.names = 1
)

colnames(counts_total_run4) <- paste0(colnames(counts_total_run4), "_2")

df <- merge(counts_total_run3, counts_total_run4, by = "row.names")
rownames(df) <- df$Row.names
df <- df[, -1]

knitr::kable(head(df))
```


## Specifying the group & batch while filterByExprs()
```{r}
counts_total <- df
d <- data.frame(
  row.names = colnames(counts_total),
  group = c(
    "MSUS", "CTRL", "MSUS", "CTRL", "MSUS",
    "CTRL", "MSUS", "CTRL", "CTRL", "MSUS",
    "MSUS", "CTRL", "MSUS", "CTRL", "MSUS",
    "CTRL", "MSUS", "CTRL", "CTRL", "MSUS"
  ),
  batch = c(
    1, 1, 1, 1, 1, 1, 1, 1, 1, 1,
    2, 2, 2, 2, 2, 2, 2, 2, 2, 2
  )
)

counts_total[is.na(counts_total)] <- 0

dds <- calcNormFactors(DGEList(counts_total), method = "TMM")

dds$samples$group <- c(
  "MSUS", "CTRL", "MSUS", "CTRL", "MSUS", "CTRL", "MSUS", "CTRL", "CTRL",
  "MSUS", "MSUS", "CTRL", "MSUS", "CTRL", "MSUS", "CTRL", "MSUS", "CTRL",
  "CTRL", "MSUS"
)

dds$samples$batch <- c(
  1, 1, 1, 1, 1, 1, 1, 1, 1, 1,
  2, 2, 2, 2, 2, 2, 2, 2, 2, 2
)

mm <- model.matrix(~ batch + group, data = d)

### specifying thebatch while filtering by expression
dds <- dds[which(filterByExpr(dds, design = mm, min.count = 20, min.prop = 0.8)), ]

# normalization by counts per million
en <- log1p(edgeR::cpm(dds))
```

## PCA plots {.tabset .tabset-pills}

### Samples colored by group and shaped by batched
```{r, warning=FALSE, message=FALSE}
plgINS::plPCA(en,
  colorBy = dds$samples$group, add.labels = FALSE,
  points.size = 12, shapeBy = paste0("Batch", dds$samples$batch)
)
```

### Samples colored by library sizes and shaped by group
```{r, warning=FALSE, message=FALSE}
plgINS::plPCA(en,
  colorBy = dds$samples$lib.size, add.labels = FALSE,
  points.size = 12, shapeBy = dds$samples$group
)
```

### Samples colored by library sizes and shaped by batches
```{r}
plgINS::plPCA(en,
  colorBy = dds$samples$lib.size, add.labels = FALSE,
  points.size = 12, shapeBy = dds$samples$batch
)
```



## Without RUVSeq:
```{r}
mm <- model.matrix(~ dds$samples$batch + dds$samples$group, data = d)
dds <- estimateDisp(dds, mm)
fit <- glmLRT(glmFit(dds, mm), coef = "dds$samples$groupMSUS")
res_without_ruvseq <- as.data.frame(topTags(fit, Inf))
knitr::kable(head(res_without_ruvseq, n = 20))
```

## Differential expression analysis with RUVSeq (SV=1) 
```{r}
re <- RUVSeq::RUVs(en,
  cIdx = row.names(en), k = 1, scIdx =
    RUVSeq::makeGroups(d$group), isLog = TRUE
)
d$SV1 <- re$W[, 1]

mm <- model.matrix(~ d$SV1 + dds$samples$batch + dds$samples$group, data = d)

dds <- estimateDisp(dds, mm)
fit <- glmLRT(glmFit(dds, mm), coef = "dds$samples$groupMSUS")
res_with_ruvseq1 <- as.data.frame(topTags(fit, Inf))
knitr::kable(head(res_with_ruvseq1, n = 20))
```


## PCA plot with 1 SV {.tabset .tabset-pills}
### Samples colored by group and shaped by batched
```{r, warning=FALSE, message=FALSE}
plgINS::plPCA(re$normalizedCounts,
  colorBy = dds$samples$group, add.labels = FALSE,
  points.size = 12, shapeBy = paste0("Batch", dds$samples$batch)
)
```

### Samples colored by library sizes and shaped by group
```{r, warning=FALSE, message=FALSE}
plgINS::plPCA(re$normalizedCounts,
  colorBy = dds$samples$lib.size, add.labels = FALSE,
  points.size = 12, shapeBy = dds$samples$group
)
```

### Samples colored by library sizes and shaped by batches
```{r, warning=FALSE, message=FALSE}
plgINS::plPCA(re$normalizedCounts,
  colorBy = dds$samples$lib.size, add.labels = FALSE,
  points.size = 12, shapeBy = dds$samples$batch
)
```


# References
```{r}
report::cite_packages(sessionInfo())
```


# SessionInfo
```{r}
devtools::session_info() %>%
  details::details()
```
